Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-03-14 bwamem.tex
the first version

\subsubsection{Seed extension} We rank the seeds by the length of the chains
the seeds belong to and then by the length of the seeds in each chain. We go
through each seed, from the best ranked to the worst, and drop the seed if it
is already ontained in an alignment found before, or use a banded
affine-gap-penalty dynamic programming to extend the seed in either direction
if the seed potentially leads to a new alignment. Up to here, we find all
the potential alignments of a query sequence.

BWA-MEM's deed extension differs from the standard seed extension in two
aspects. Firstly, suppose at a certain extension step we come to reference
position $x$ with the best extension score achieved at query position $y$.
BWA-MEM will stop extension if the score at $(x,y)$ is smaller than the best
extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
$p_{\rm gapExt}$ is the gap extension penalty. This heuristic avoids extending
through a poorly aligned region with good flanking alignment. It is similar
to the X-dropoff heuristic in BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in
one of the reference or the query sequences.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (top panel) or as paired end (bottom panel). GEM is configured to
allow gaps and output suboptimal alignments (option --e5 --m5 --s1). Its
mapping quality is estimated with a BWA-like algorithm. Other mappers are run
essentially with the default setting except specifying the insert size
distribution. CPU timeU is shown in the parantheses.} \end{figure}\label{fig:eval}.

@@

\subsubsection{Pairing} Given the $i$-th hit for the first read and $j$-th hit
for the second, BWA-MEM computes their distance $d_{ij}$ if the two hits are in
the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
pair $(i,j)$ as $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$ and
$S_j$ are the SW score of the two reads in the pair, respectively, $a$ is the
matching score and $P(d)$ gives the probability of observing an insert size
larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
relate SW score to a probabilistic alignment. $U$ is the unpairing penalty,
which controls whether pairing is preferred: if $d_{ij}$ is small enough (such
that $-a\log_4 P(d_{ij})<U$), BWA-MEM prefers to pair the two ends; otherwise
it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
maximizes $S_{ij}$ as the final alignment for both ends.

This pairing strategy jointly considers the alignment scores of both ends,
the insert size and the possibility that they may come from a chimeric read
pair. It prefers pairing the two ends but does not aggressively force two ends
into a read pair.

@@

\section{Results}

We evaluated the performance of BWA-MEM on simulated data together with
GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
(http://novocraft.com) (Figure~\ref{fig:eval}). We were unable to obtain high
accuracy for paired-end data with GEM, so we did not show it result. In this
evaluation, NovoAlign is the most accurate. BWA-MEM is only second to NovoAlign
in accuracy and among the fastest in speed.

\section{Conclusion}

BWA-MEM is a fast and accurate alignment algorithm for sequence reads. It is
one of the few aligners that work with both short reads and long sequences up
to a few megabases.

>>>>

\subsubsection{Seed extension} We rank the seeds by the length of the chains
the seeds belong to and then by the length of the seeds in each chain. We start
from the highest ranked seed and for each seed, we drop the seed if it is
already contained in an alignment found before, or extend with a
affine-gap-penalty dynamic programming (DP) otherwise. We will find all the
potential alignments of a query sequence after this step.

BWA-MEM's seed extension differs from the standard seed extension in two
aspects. Firstly, suppose at a certain extension step we come to reference
position $x$ with the best extension score achieved at query position $y$.
BWA-MEM will stop extension if the score at $(x,y)$ is smaller than the best
extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary value. This heuristic avoids extending
through a poorly aligned region with good flanking alignment. It is similar
to the X-dropoff heuristic in BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in
one of the reference or the query sequences.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (SE; top panel) or as paired end (PE; bottom panel). GEM is configured to
allow up to 5 gaps and to output suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its
mapping quality is estimated with a BWA-like algorithm. Other mappers are run
essentially with the default setting except specifying the insert size
distribution. The run time on a single CPU core is shown in the parantheses.}\label{fig:eval}
\end{figure}.

@@

\subsubsection{Pairing} Given the $i$-th hit for the first read and $j$-th hit
for the second, BWA-MEM computes their distance $d_{ij}$ if the two hits are in
the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
pair $(i,j)$ as $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$ and
$S_j$ are the SW score of the two reads in the pair, respectively, $a$ is the
matching score and $P(d)$ gives the probability of observing an insert size
larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
relate SW score to a probabilistic model. $U$ is the chimeric penalty,
which controls preference of pairing: if $d_{ij}$ is small enough such
that $-a\log_4 P(d_{ij})<U$, BWA-MEM prefers to pair the two ends; otherwise
it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
maximizes $S_{ij}$ as the final alignment for both ends.
This pairing strategy jointly considers the alignment scores of both ends,
the insert size and the possibility of chimeric read pairs.

@@

\section{Results and Discussions}

We evaluated the performance of BWA-MEM on simulated data together with
GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
(http://novocraft.com) (Figure~\ref{fig:eval}). On accuracy, NovoAlign
is the most accurate. BWA-MEM is close to NovoAlign for PE reads
and is comparable to GEM and Cushaw2 for SE. On speed, BWA-MEM is similar to
GEM and Bowtie2 for this data set, but is about 6 times as fast as Bowtie2
for a 650bp long-read data set.

To evaluate the performance for even long query sequences, we aligned the {\it E. coli}
K-12 MG1655 substrain (AC:NC\_000913) against the 536 strain (AC:NC\_008253) with 
both BWA-MEM and nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds, giving
105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of which 102,241
overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short regions
with very high divergence. It is unclear which aligner is better. Although
Nucmer is faster in the evaluation, only BWA-MEM scales to mammalian genomes.

BWA-MEM is a fast and accurate alignment algorithm for sequence reads and is
one of the few aligners that work efficiently with both short reads and long sequences up
to a few megabases. It is still possible to greatly accelerate BWA-MEM for long
sequences by using SSE2-powered banded DP and by restricting DP to regions not
covered by a long seed.
